### 2014-12-01 revision/Paper_v3.tex
JVs changes for reviewers
JVs changes for revised manuscript

The rule-based regulatory strategy approximated the behavior of classical allosteric activation and inhibition mechanisms (Fig. \ref{fig-kinetics-simulations}). 
We considered the enzyme catalyzed conversion of substrate S to a product P, where the overall reaction rate was modeled as the product of a Michaelis-Menten term and an effective allosteric control variable reflecting the particular regulatory interaction. 
We first explored feed-forward substrate activation of enzyme activity (for both positive and negative cooperativity). 
Consistent with classical data, the rule-based strategy predicted a sigmoidal relationship between substrate abundance and reaction rate as a function of the cooperativity parameter (Fig. \ref{fig-kinetics-simulations}A). 
For cooperativity parameters less than unity, increased substrate abundance \textit{decreased} the reaction rate. 
This was consistent with the idea that substrate binding \textit{decreased} at regulatory sites, which negatively impacted substrate binding at the active site. 
On the other hand, as the cooperativity parameter increased past unity, the rate of conversion of substrate S to product P by enzyme $E$ approached a step function. 
In the presence of an inhibitor, the rule-based strategy predicted non-competitive like behavior as a function of the cooperativity parameter (Fig. \ref{fig-kinetics-simulations}B). 
When the control gain parameter, $\kappa_{ij}$ in Eqn. \eqref{eqn:control-factor}, was greater than unity, the inhibitory force was directly proportional to the cooperativity parameter, $\eta$ in Eqn. \eqref{eqn:control-factor}. 
Thus, as the cooperativity parameter increased, the maximum reaction rate decreased (Fig. \ref{fig-kinetics-simulations}B). 
Interestingly, our rule-based approach was unable to directly simulate competitive inhibition of enzyme activity. 
Taken together, the rule-based strategy captured classical regulatory patterns for both enzyme activation and inhibition. 
Thus, we are able to model complex kinetic phenomena such as ultrasensitivity, despite an effective description of reaction kinetics.

@@

\subsection*{Estimating parameters and effective allosteric regulatory structures.}
A critical challenge for any dynamic model is the estimation of kinetic parameters. 
For metabolic processes, there is also the added challenge of identifying the regulation and control structures that manage metabolism. 
Of course, these issues are not independent; any description of enzyme activity regulation will be a function of system state, which in turn depends upon the kinetic parameters. 
For cell-free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
We explored this linkage by estimating model parameters from synthetic data using both network structures. 
We generated noise corrupted synthetic measurements of the substrate S, intermediate M$_{5}$ and end product P$_1$ approximately every 20 min using
network A. We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm optimization (PSO), starting from random initial parameter guesses. The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig. \ref{fig-parameter-fit}). 
PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Fig. \ref{fig-parameter-fit}A and B). 
However, with control mismatch (network B simulated with network A parameters), model simulations were not consistent with the synthetic data  (Fig. \ref{fig-parameter-fit}C and D). 
Taken together, these results suggested that we could perhaps simultaneously estimate both parameters and network control architectures, as incorrect control structures would be manifest as poor model fits.

@@

@@

The allosteric regulation term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
For each enzyme, we used a rule-based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
If an enzyme was activated by $m$ metabolites, we modeled this activation as:
\begin{equation}
	v_{j} = \max\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right)
\end{equation}where $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$ was a regulatory transfer function that calculated the influence of metabolite $i$ on the activity of enzyme $j$. 
Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
\begin{equation}
	v_{j} = 1 - \max\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right)
\end{equation}Lastly, if an enzyme had both $m$ activating and $n$ inhibitory factors, we modeled the regulatory term as:
\begin{equation}
	v_{j} = \min\left(u_{j},d_{j}\right)
\end{equation}where:
\begin{eqnarray}
	u_{j} &=& \max_{j^{+}}\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right) \\
	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{nj}\left(\mathcal{Z}\right)\right)
\end{eqnarray}
The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating and inhibitory factors for enzyme $j$. 
If an enzyme had no allosteric factors, we set $v_{j} = 1$.
There are many possible functional forms for $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$. 
However, in this study, each individual transfer function took the form:
\begin{equation}\label{eqn:control-factor}
	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}
\end{equation}where $\mathcal{Z}_{j}$ denotes the abundance of the $j$ factor (e.g., metabolite abundance), and $\kappa_{ij}$ and $\eta$ are control parameters. 
The $\kappa_{ij}$ parameter represents a species gain parameter, while $\eta$ is a cooperativity parameter (similar to a Hill coefficient). 
In the case $\eta>1$, the allosteric interaction displays positive cooperativity. For $\eta<1$, the interaction is negatively cooperative. Finally, if $\eta=1$, the interaction displays no cooperativity. 
The effect of different values of $\eta$ on reaction rate can be seen in Figure \ref{fig-kinetics-simulations}.
The model equations were encoded using the Octave programming language and solved using the LSODE routine in Octave (v 3.8.1; www.octave.org).

\subsection*{Estimation of model parameters and structures from synthetic experimental data.}
Model parameters were estimated by minimizing the difference between simulations and synthetic experimental data (squared residual):
\begin{equation}\label{eqn:objective-function}
	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated 
value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is respect to
time, while the inner summation is with respect to state. We approximated a realistic model identification scenario, assuming noisy experimental data, 
limited sampling resolution (approximately 20 minutes per sample) and a limited number of measurable metabolites.

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic using particle swarm optimization (PSO). 
Synthetic experimental data was generated from a hypothetical parameter set using Network A, 
where substrate $S$, end product P$_{1}$ and intermediate M$_5$ were sampled approximately every 20 minutes. 
For cases $\textbf{A,B}$ 20 particles were initialized with randomized parameters and allowed to search for 300 iterations. 
\textbf{A,B}: PSO estimated an ensemble of 20 parameters sets consistent with the synthetic experimental data assuming the correct
enzymatic and control connectivity starting from randomized initial parameters.
\textbf{C,D}: In the presence of control mismatch (Network B control policy simulated with Network A kinetic parameters) 
the ensemble of models did not describe the synthetic data. 
}\label{fig-parameter-fit}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth,height=0.6\textheight]{./figs/Figure-8-Flux.pdf}
\caption{Metabolic flux and control variables as a function of network type and particle index at t = 100 min.
The control variables governing $E_{1},E_{3}$ and $E_{6}$ activity and the scaled metabolic flux and were calculated for the positively, 
uniformly and negatively biased particle swarms (N = 100). The particles from each swarm were sorted based upon simulation error (low to high error). 
\textbf{A}: Model performance for the positively biased particle swarm as a function of particle index.
\textbf{B}: Model performance for the uniformly biased particle swarm as a function of particle index.
\textbf{C}: Model performance for the negatively biased particle swarm as a function of particle index. 
Models with significant control mismatch showed distinct control and flux patterns versus those models with the correct or closely related control policies.
In particular, models with the correct control policy showed stronger inhibition of $E_{1}$ activity, leading to decreased flux from S$\rightarrow$P$_{1}$.
Conversely, models with significant mismatch had increased $E_{1}$ activity, leading to an altered flux distribution. 
This is especially apparent in the negatively biased particle swarm. 
}\label{fig-flux-pattern}
\end{figure}

>>>>

The rule-based regulatory strategy approximated the behavior of classical allosteric activation and inhibition mechanisms (Fig. \ref{fig-kinetics-simulations}). 
We considered the enzyme catalyzed conversion of substrate S to a product P, where the overall reaction rate was modeled as the product of a Michaelis-Menten term and an effective allosteric control variable reflecting the particular regulatory interaction. 
We first explored feed-forward substrate activation of enzyme activity (for both positive and negative cooperativity). 
Consistent with classical data, the rule-based strategy predicted a sigmoidal relationship between substrate abundance and reaction rate as a function of the cooperativity parameter (Fig. \ref{fig-kinetics-simulations}A). 
For cooperativity parameters less than unity, increased substrate abundance \textit{decreased} the maximum reaction rate. 
This was consistent with the idea that substrate binding \textit{decreased} at regulatory sites, which negatively impacted substrate binding at the active site. 
On the other hand, as the cooperativity parameter increased past unity, the rate of conversion of substrate S to product P by enzyme $E$ approached a step function. 
In the presence of an inhibitor, the rule-based strategy predicted non-competitive like behavior as a function of the cooperativity parameter (Fig. \ref{fig-kinetics-simulations}B). 
When the control gain parameter, $\kappa_{ij}$ in Eqn. \eqref{eqn:control-factor}, was greater than unity, the inhibitory force was directly proportional to the cooperativity parameter, $\eta$ in Eqn. \eqref{eqn:control-factor}. 
Thus, as the cooperativity parameter increased, the maximum reaction rate decreased (Fig. \ref{fig-kinetics-simulations}B). 
Interestingly, our rule-based approach was unable to directly simulate competitive inhibition of enzyme activity. 
Taken together, the rule-based strategy captured classical regulatory patterns for both enzyme activation and inhibition. 
Thus, we are able to model complex kinetic phenomena such as ultrasensitivity, despite an effective description of reaction kinetics.

@@

\subsection*{Estimating parameters and effective allosteric regulatory structures.}
A critical challenge for any dynamic model is the estimation of kinetic parameters. 
For metabolic processes, there is also the added challenge of identifying the regulation and control structures that manage metabolism. 
Of course, these issues are not independent; any description of enzyme activity regulation will be a function of system state, which in turn depends upon the kinetic parameters. 
For cell free systems, regulated gene expression has been removed, however, enzyme activity regulation is still operational. 
We explored this linkage by estimating model parameters from synthetic data using both network structures. 
We generated synthetic measurements of the substrate S, intermediate M$_{5}$ and end product P$_1$ approximately every 20 min using
network A. This data set is similar to published cell free studies both in terms of network coverage, and sampling frequency \citep{Jewett:2008aa}. 
We then generated an ensemble of model parameter estimates by minimizing the difference between model simulations and the synthetic data using particle swarm optimization (PSO), starting from random initial parameter guesses. The estimation of kinetic parameters was sensitive to the choice of regulatory structure (Fig. \ref{fig-parameter-fit}). 
PSO identified an ensemble of parameters that bracketed the mean of the synthetic measurements in less than 1000 iterations when the control structure was correct (Fig. \ref{fig-parameter-fit}A and B). 
However, with control mismatch (network B simulated with network A parameters), model simulations were not consistent with the synthetic data  (Fig. \ref{fig-parameter-fit}C and D). 
Taken together, these results suggested that we could perhaps simultaneously estimate both parameters and network control architectures, as incorrect control structures would be manifest as poor model fits.

@@

Lastly, the choice of max/min integration rules or the particular form of the transfer functions could be generalized to include other rule types and functions. 
Theoretically, an integration rule is a function whose domain is a set of transfer function inputs, and whose range is $v\in[0,1]$.
Thus, integration rules other than max/min could be used, such as the mean or the product, assuming the range of the transfer functions is always $f\in[0,1]$.
Alternative integration rules such as the mean might have different properties which could influence model identification or performance. 
For example, a mean integration rule would be differentiable, which allows derivative-based optimization approaches to be used. 
The particular form of the transfer function could also be explored. We choose a Hill-like function because of its 
prominence in the systems and synthetic biology community. However, the only mathematical requirement for a transfer function is that it map a non-negative continuous or categorical variable into the 
range $f\in[0,1]$. Thus, many types of transfer functions are possible.

@@

The allosteric regulation term $v_{j}$ depended upon the combination of factors which influenced the activity of enzyme $i$.
For each enzyme, we used a rule-based approach to select from competing control factors (Fig. \ref{fig-control-schematic}). 
If an enzyme was activated by $m$ metabolites, we modeled this activation as:
\begin{equation}
	v_{j} = \max\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right)
\end{equation}where $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$ was a regulatory transfer function that calculated the influence of metabolite $i$ on the activity of enzyme $j$. 
Conversely, if enzyme activity was inhibited by a $m$ metabolites, we modeling this inhibition as:
\begin{equation}
	v_{j} = 1 - \max\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right)
\end{equation}Lastly, if an enzyme had both $m$ activating and $n$ inhibitory factors, we modeled the regulatory term as:
\begin{equation}
	v_{j} = \min\left(u_{j},d_{j}\right)
\end{equation}where:
\begin{eqnarray}
	u_{j} &=& \max_{j^{+}}\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{mj}\left(\mathcal{Z}\right)\right) \\
	d_{j} &=& 1 - \max_{j^{-}}\left(f_{1j}\left(\mathcal{Z}\right),\hdots,f_{nj}\left(\mathcal{Z}\right)\right)
\end{eqnarray}
The quantities $j^{+}$ and $j^{-}$ denoted the sets of activating and inhibitory factors for enzyme $j$. 
If an enzyme had no allosteric factors, we set $v_{j} = 1$.
There are many possible functional forms for $0\leq f_{ij}\left(\mathcal{Z}\right)\leq 1$. 
However, in this study, each individual transfer function took the form:
\begin{equation}\label{eqn:control-factor}
	f_{i}\left(\mathbf{x}\right) = \frac{\kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}{1 + \kappa_{ij}^{\eta}\mathcal{Z}_{j}^{\eta}}
\end{equation}where $\mathcal{Z}_{j}$ denotes the abundance of the $j$ factor (e.g., metabolite abundance), and $\kappa_{ij}$ and $\eta$ are control parameters. 
The $\kappa_{ij}$ parameter represents a species gain parameter, while $\eta$ is a cooperativity parameter (similar to a Hill coefficient). 
In the case $\eta>1$, the allosteric interaction displays positive cooperativity. For $\eta<1$, the interaction is negatively cooperative. Finally, if $\eta=1$, the interaction displays no cooperativity. 
The effect of different values of $\eta$ on reaction rate can be seen in Figure \ref{fig-kinetics-simulations}.
The model equations were encoded using the Octave programming language and solved using the LSODE routine in Octave (v 3.8.1; www.octave.org). 
In some cases, metabolic fluxes (or other quantities) were scaled according to:
\begin{equation}
	\hat{r}_{j}\left(t = \tau\right) = \left(\frac{r_{j} - \min\mathbf{r}}{\max\mathbf{r} - \min\mathbf{r}}\right)\Bigr|_{t = \tau}
\end{equation}where $0\leq\hat{r}_{j}\left(t = \tau\right)\leq{1}$ denotes the scaled value for flux j evaluated at time $\tau$. 
We have used this scaling in a variety of other contexts \citep{Song:2010ab,Lequieu:2011aa}.

\subsection*{Estimation of model parameters and structures from synthetic experimental data.}
Model parameters were estimated by minimizing the difference between simulations and synthetic experimental data (squared residual):
\begin{equation}\label{eqn:objective-function}
	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated 
value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is respect to
time, while the inner summation is with respect to state. We approximated a realistic model identification scenario, assuming noisy experimental data, 
limited sampling resolution (approximately 20 minutes per sample) and a limited number of measurable metabolites. We assumed a constant coefficient of variation of 10\% for the synthetic data set.

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic using particle swarm optimization (PSO). 
Synthetic experimental data was generated from a hypothetical parameter set using Network A, 
where substrate $S$, end product P$_{1}$ and intermediate M$_5$ were sampled approximately every 20 minutes. 
For cases $\textbf{A,B}$ 20 particles were initialized with randomized parameters and allowed to search for 300 iterations. 
\textbf{A,B}: PSO estimated an ensemble of 20 parameters sets consistent with the synthetic experimental data assuming the correct
enzymatic and control connectivity starting from randomized initial parameters.
\textbf{C,D}: In the presence of control mismatch (Network B control policy simulated with Network A kinetic parameters) 
the ensemble of models did not describe the synthetic data. The synthetic data plotted here was unperturbed by noise. 
However, we assumed a coefficient of variation of 10\% for the synthetic data during parameter estimation.
}\label{fig-parameter-fit}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth,height=0.6\textheight]{./figs/Figure-8-Flux.pdf}
\caption{Metabolic flux and control variables as a function of network type and particle index at t = 100 min.
The control variables governing $E_{1},E_{3}$ and $E_{6}$ activity and the scaled metabolic flux were calculated for the positively (top), 
uniformly (middle) and negatively (bottom) biased particle swarms (N = 100). 
Blue denotes a low value, while red denotes a high value for the respective quantity being plotted.
The particles from each swarm were sorted based upon simulation error (low to high error). 
\textbf{A}: Model performance for the positively biased particle swarm as a function of particle index.
\textbf{B}: Model performance for the uniformly biased particle swarm as a function of particle index.
\textbf{C}: Model performance for the negatively biased particle swarm as a function of particle index. 
Models with significant control mismatch showed distinct control and flux patterns versus those models with the correct or closely related control policies.
In particular, models with the correct control policy showed stronger inhibition of $E_{1}$ activity, leading to decreased flux from S$\rightarrow$P$_{1}$.
Conversely, models with significant mismatch had increased $E_{1}$ activity, leading to an altered flux distribution. 
This is especially apparent in the negatively biased particle swarm. 
}\label{fig-flux-pattern}
\end{figure}
